Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-03-10 src/tex/ms.tex
190521 edits

It is difficult to unequivocally identify a specific feature of the GW170818 data that leads to this posterior structure.
However, it appears to be related to this event's support for precession, in conjunction with its uncommonly definite measurement of the polarization and spin angles \cite{Varma:2021csh}.
The relevance of precession is evident from the posterior on $\chi_p$ (Fig.~\ref{fig:hist_GW170818}): allowing for $\kappa \neq 0$ leads to reduced support for precession.
We can understand this in reference to Fig.~\ref{fig:birefringence}: if only a short portion of the signal is observed, then the frequency-dependent signal enhancement or dampening due to birefringence can mimic the time-dependent amplitude modulation in a precession cycle.
Therefore, under these circumstances, similar morphologies can be obtained by setting $\chi_p > 0$ or $\kappa > 0$, as long as the distance and inclination can also be adjusted accordingly.

@@

\begin{figure}
    \script{corner_GW170818.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW170818.pdf}
    \caption{GW170818 posterior on $\kappa$, luminosity distance $d_L$ and inclination $\cos\iota$ from our birefringence analysis (blue), compared to the GR result (orange).}
    \label{fig:corner_GW170818}
\end{figure}

\begin{figure}
    \script{hist_GW170818.py}
    \includegraphics[width=\columnwidth]{figures/hist_GW170818.pdf}
    \caption{
        GW170818 posterior on the precessing spin $\chi_p$, as obtained by allowing birefringence (blue) and not (orange). \mi{TODO (for Max): maybe make this an inset to Fig.~\ref{fig:corner_GW170818}.}
    }
    \label{fig:hist_GW170818}
\end{figure}

@@

GW190521 is the most extreme representative of a class of events with bimodal $\kappa$ posteriors (Table \ref{tab:bimodal_events_mass}).
This appears to arise out of a confluence of factors related to this event's high mass and support for precession: this being the most massive event in our set, the signal is only in-band for a very short time before merger, meaning that spin effects can more easily be approximated by frequency-dependent birefringent.
In particular, a precession cycle can more easily by approximated by a monotonic amplitude modulation.
This effect gives rise to a similar degeneracy between $\kappa$ and $\chi_p$ as outlined for GW170818, wherein the prior volume favored configurations with $|\kappa| >0$.
However, in the case of GW190521, the polarization measurement at the detectors is not accurate enough to fully disambiguate between right- and left-handed states (as implied by the bimodal $\cos\iota$ posterior obtained assuming \ac{GR} in Fig.~\ref{fig:corner_GW190521}).
As a result, the $\kappa$ posterior is bimodal for GW190521.

The degeneracy between $\kappa$ and $\iota$ cannot be broken by the frequency-dependent birefringence model, as shown in Fig.~\ref{fig:corner_GW190521}.
The effect of birefringence at different frequencies within the range is similar, so the degeneracy cannot be broken.
Therefore, the narrow frequency range can be a possible reason why the $\kappa$ distribution of GW190521 is bimodal.
Note that $\kappa=0$ is still within the $90\%$ confidence interval, with a non-negligible probability.
The bimodal $\kappa$ distribution only means that the degeneracy between $\kappa$ and $\iota$ cannot be broken in this case.  \wf{You should still explain \emph{why} the model seems to prefer $\left| \kappa \right| > 0$.  I \emph{think} the reason here is the distance prior---there is a lot more ``phase space'' at large distances that can be accessed by allowing $\kappa$ to be non-zero and choosing a corresponding inclination so that the amplification effects of birefringence are compensated for by the larger distance compared to the pure-GR run.  Does this make sense to you?  If so, can you write it up in this paragraph?}

\begin{figure}[h]
    \script{corner_GW190521.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW190521.pdf}
    \caption{
        GW190521 posterior on $\kappa$, luminosity distance $d_L$ and inclination $\cos\iota$ from our frequency-dependent (blue) and frequency-independent (orange) birefringence analysis, compared to the GR result (green).
        % The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW190521.
        % Colors in the plot represent the \ac{PE} result with \ac{GR} and the frequency-dependent birefringence, respectively.
        % The 2D contours correspond to the $39.35\%$ ($1\sigma$) and $90\%$ credible levels.
        % This plot shows that frequency-dependent birefringence cannot break the degeneracy between $\kappa$ and $\iota$.
    }
    \label{fig:corner_GW190521}
\end{figure}

@@

\begin{figure}
    \script{corner_GW150914.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW150914.pdf}
    \caption{
        GW150914 posterior on $\kappa$, luminosity distance $d_L$ and inclination $\cos\iota$ from our frequency-dependent (blue) and frequency-independent (orange) birefringence analysis, compared to the GR result (green).
        % The posterior of $\kappa$, luminosity distance $d_L$ and $\cos{\iota}$ for GW150914.
        % Colors in the plot represent the \ac{PE} result with \ac{GR}, frequency-independent and dependent birefringence, respectively.
        % The 2D contours correspond to the $39.35\%$ ($1\sigma$) and $90\%$ credible levels.
        % Note that there is no posterior of $\kappa$ for the \ac{PE} result with \ac{GR}, as the result is based on GR, which does not suggest GW amplitude birefringence.
        % This plot shows that frequency-independent birefringence creates a degeneracy between $\kappa$ and $\iota$, while frequency-dependent birefringence can break it.
    }
    \label{fig:corner_GW150914}
\end{figure}

@@

\begin{figure}
    \script{corner_GW200129.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW200129.pdf}
    \caption{
        GW200129 posterior on $\kappa$, luminosity distance $d_L$ and inclination $\cos{\iota}$, including different sets of detectors in the analysis per the legend.
        The main run with all three detectors (HLV, filled blue), shows a preference for $\kappa < 0$, as in Fig.~\ref{fig:violin_kappa}; this preference is more pronounced for two-detector runs that include Virgo (HV and LV, orange and green); however, it disappears if we remove Virgo (HL, dashed black).
        The 2D contours correspond to the $90\%$ credible level.
    }
    \label{fig:corner_GW200129}
\end{figure}

>>>>

It is difficult to unequivocally identify a specific feature of the GW170818 data that leads to this posterior structure.
However, it appears to be related to this event's support for precession, in conjunction with its uncommonly definite measurement of the polarization and spin angles \cite{Varma:2021csh}.
The relevance of precession is evident from the posterior on $\chi_p$ (Fig.~\ref{fig:corner_GW170818}): allowing for $\kappa \neq 0$ leads to reduced support for precession.
We can understand this in reference to Fig.~\ref{fig:birefringence}: if only a short portion of the signal is observed, then the frequency-dependent signal enhancement or dampening due to birefringence can mimic the time-dependent amplitude modulation in a precession cycle.
Therefore, under these circumstances, similar morphologies can be obtained by setting $\chi_p > 0$ or $\kappa > 0$, as long as the distance and inclination can also be adjusted accordingly.

@@

\begin{figure}
    \script{corner_GW170818.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW170818.pdf}
    \caption{
        GW170818 posterior on $\kappa$, luminosity distance $d_L$ and inclination $\cos\iota$ from our birefringence analysis (blue), compared to the GR result (orange).
        The marginalized posterior on $\chi_p$ is shown in the top right panel.
    }
    \label{fig:corner_GW170818}
\end{figure}

@@

GW190521 is the most extreme representative of a class of events with bimodal $\kappa$ posteriors (Table \ref{tab:bimodal_events_mass}).
This appears to arise out of a confluence of factors related to this event's high mass and support for precession: GW190521 being the most massive event in our set, its signal is only in-band for a very short time before merger, meaning that frequency-dependent birefringence can more easily resemble spin effects.
In particular, a precession cycle can be approximated by a monotonic amplitude modulation for a short time.
This effect gives rise to a similar degeneracy between $\kappa$ and $\chi_p$ as outlined for GW170818, wherein the prior volume favored configurations with $|\kappa| >0$.
However, in the case of GW190521, the polarization measurement at the detectors is not accurate enough to fully disambiguate between right- and left-handed states (as implied by the bimodal $\cos\iota$ posterior obtained assuming \ac{GR} in Fig.~\ref{fig:corner_GW190521}).
As a result, the GW190521 posterior does not favor $\kappa > 0$ exclusively, but supports both $\kappa > 0$ and $\kappa < 0$, leading to a bimodal posterior.

\begin{figure}[h]
    \script{corner_GW190521.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW190521.pdf}
    \caption{
        GW190521 posterior on $\kappa$, luminosity distance $d_L$ and inclination $\cos\iota$ from our frequency-dependent (blue), compared to the GR result (green).
        The marginalized posterior on $\chi_p$ is shown in the top right panel.
    }
    \label{fig:corner_GW190521}
\end{figure}

@@

\begin{figure}
    \script{corner_GW150914.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW150914.pdf}
    \caption{
        GW150914 posterior on $\kappa$, luminosity distance $d_L$ and inclination $\cos\iota$ from our frequency-dependent (blue), compared to the GR result (green).
        The marginalized posterior on $\chi_p$ is shown in the top right panel.
    }
    \label{fig:corner_GW150914}
\end{figure}

@@

\begin{figure}
    \script{corner_GW200129.py}
    \includegraphics[width=\columnwidth]{figures/corner_GW200129.pdf}
    \caption{
        GW200129 posterior on $\kappa$, luminosity distance $d_L$ and inclination $\cos{\iota}$, including different sets of detectors in the analysis per the legend.
        The marginalized posterior on $\chi_p$ is shown in the top right panel.
        The main run with all three detectors (HLV, filled blue), shows a preference for $\kappa < 0$, as in Fig.~\ref{fig:violin_kappa}; this preference is more pronounced for two-detector runs that include Virgo (HV and LV, orange and green); however, it disappears if we remove Virgo (HL, dashed black).
        The 2D contours correspond to the $90\%$ credible level.
    }
    \label{fig:corner_GW200129}
\end{figure}
